Exploring Crete's Macroecology based on its soil Microbial Interactome

### 2024-02-26 manuscript.tex
move to cite from citep

Soil ecosystems are the cornstone of terrestrial habitats, biodiversity and henceforth human activities.
Complex interactions of multiple systems, nutrient cycling, atmosphere, climate, biogeochemical cycles and biodiversity 
across all forms of life, fauna, flora, bacteria, archaea, fungi, viruses. 
Bacteria and archaea are considered major drivers for the functionality of soil.
They infuence and are infuenced by their environments and their community structure 
is defies their macroscopic functionality.
Global soil microbiome studies \citep{Delgado-Baquerizo2018} and \citep{Bahram2018} and the biogeography \citep{Martiny2006} \citep{Guerra2020}, \citep{Labouyrie2023} yet there are significant blind spots \citep{Guerra2020}. The most resolute study is \citep{Karimi2020} which exemplified the 
remarkable complexity of soil bacterial communuties and the requirement of dense samplings and isolated systems.

Island Sampling Day project and the 
Genome Standards Consortium \citep{Field2011} "put standards into action" and the 
18th workshop. 
The choice of an island
Islands are nature's labs \citep{Whittaker2017} and in soil microbes \citep{Li2020} and fungi \citep{Delavaux2021}
Ultimately islands can set the ground to represent ecosystems with data and complex interactions \citep{Davies2016}.
A large scale study in the confined space of the island of Crete, a miniature continent.

The island of Crete, Greece has been studied since the classical times for its' fauna \citep{Sidiropoulos_Polymeni_Legakis_2017,Anastasiou2018Tenebrionid}, flora \citep{Krimbas_2005} and ecosystems \citep{Grove1993}.
Crete is home to the only endemic mammal of Greece, the Cretan shrew (\textit{Crocidura zimmermanni}), more than 350 endemic arhropods, \# endemic plants among them a tree \textit{Zelkova abelicea}.
Crete is a continental island, fifth largest island of the mediterranean, a mediterranean biodiversity hotspot, model for desertification, climate change \citep{Kougioumoutzis2020} and habitat loss \citep{ISPIKOUDIS1993259},
elevation gradient \citep{trigas2013, FAZAN2017}, grasing impacts on ecosystems \citep{JouffroyBapicot2016} and evolutionary model \citep{POULAKAKIS2002}.
Yet the terrestrial microbial diversity has been unexplored.

Has been sampled from major worldwide projects of microbiome studies \citep{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}. Soil fungi \citep{Mikryukov2023, Davison2021, Tedersoo2021} and other soil eukaryotes \citep{Aslani2022}

\citep{Vogiatzakis2008_crete}
\citep{Vogiatzakis2016}

\citep{Sfenthourakis2017}
\citep{Vogiatzakis2008_med}

Integration of prior knowledge and 
Genome Standards Consortium \citep{Field2011}, FAIR data \citep{Wilkinson2016}, unrestricted \citep{Amann2019},
rich metadata and platforms of hosting open data like European Soil Data Centre \citep{Panagos2022}

Interactomes as a driver of community composition. Microbial interactions challenges \citep{Faust2021} cohesion?

@@

the data for this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB21776
ENA API \citep{Yuan2023}

@@

Copernicus CORINE Land Cover \citep{CLC2023}
WorldClim 2.0 \citep{Fick2017}
World Database on Protected Areas (WDPA) \citep{Hanson2022}
Function annotation - FAPROTAX \citep{Louca2016}
Crete geology from IGME
Aridity
%NDVI
Taxonomy Silva 138 \citep{Quast2012}

\subsection{Tools}\label{tools}
dada2 for ASV inferrence \citep{Callahan2016}
PEMA for OTU inferrence \citep{Zafeiropoulos2020}
ape R packege \citep{Paradis2004}
SRS R package \citep{Beule2020}
UMAP ordination \citep{mcinnes2018umap-software}
Numerical ecology analyses with vegan R package \citep{vegan}

Python 3.11.4
Julia Version 1.9.3 \citep{Julia-2017}
R version 4.3.2 \citep{rcoreteam}
Network inferrence using FlashWeave 0.19.2 \citep{Tackmann2019}
Network analysis in igraph R package \citep{Csardi2006}
Differential abundance analysis with ANCOM-BC R package \citep{Lin2020,Lin2023}
Spatial analysis with the sf and stars R packages \citep{Pebesma2023}
U-CIE R package for coloring 3 dimentional data \citep{Koutrouli2022}
ggplot2 \citep{ggplot22016}
pheatmap \citep{Kolde2019}

Computations were performed on HPC infrastrure of HCMR \citep{Zafeiropoulos2021}.

@@

Networks specialists generalists \citep{Barberan2012}

Elevation gradients of biodiversity are known since Humboldt's work \citep{Rahbek2019} 
yet these patterns remain elusive regarding the soil microbiome \citep{Looby2020, Siles2023}.
Mostly because it's difficult to isolate other cofounding effects \citep{Nottingham2018}.

@@

Enumeratic all biodivarsity \citep{Anthony2023} and decipher the biogeochemical 
processes and interactions of fauna, flora and microbes \citep{Fry2019, Crowther2019,GRANDY201640,Delgado-Baquerizo2020}

Global hotspots \citep{Guerra2022}
and soil conservation instead of specific species \citep{Guerra2021}
and policy \citep{KONINGER2022} across countries \citep{Putten2023}

@@

Mountain peaks soil microbes \citep{Adamczyk2019}
Even though amplicon studies in soil should be interpeted with caution \citep{alteio2021} they 
can act as early warning signals towards public health conserns \citep{Banerjee2023}.
Some Conclusions here.

>>>>

Soil ecosystems are the cornstone of terrestrial habitats, biodiversity and henceforth human activities.
Complex interactions of multiple systems, nutrient cycling, atmosphere, climate, biogeochemical cycles and biodiversity 
across all forms of life, fauna, flora, bacteria, archaea, fungi, viruses. 
Bacteria and archaea are considered major drivers for the functionality of soil.
They infuence and are infuenced by their environments and their community structure 
is defies their macroscopic functionality.
Global soil microbiome studies \cite{Delgado-Baquerizo2018} and \cite{Bahram2018} and the biogeography \cite{Martiny2006} \cite{Guerra2020}, \cite{Labouyrie2023} yet there are significant blind spots \cite{Guerra2020}. The most resolute study is \cite{Karimi2020} which exemplified the 
remarkable complexity of soil bacterial communuties and the requirement of dense samplings and isolated systems.

Island Sampling Day project and the 
Genome Standards Consortium \cite{Field2011} "put standards into action" and the 
18th workshop. 
The choice of an island
Islands are nature's labs \cite{Whittaker2017} and in soil microbes \cite{Li2020} and fungi \cite{Delavaux2021}
Ultimately islands can set the ground to represent ecosystems with data and complex interactions \cite{Davies2016}.
A large scale study in the confined space of the island of Crete, a miniature continent.

The island of Crete, Greece has been studied since the classical times for its' fauna \cite{Sidiropoulos_Polymeni_Legakis_2017,Anastasiou2018Tenebrionid}, flora \cite{Krimbas_2005} and ecosystems \cite{Grove1993}.
Crete is home to the only endemic mammal of Greece, the Cretan shrew (\textit{Crocidura zimmermanni}), more than 350 endemic arhropods, \# endemic plants among them a tree \textit{Zelkova abelicea}.
Crete is a continental island, fifth largest island of the mediterranean, a mediterranean biodiversity hotspot, model for desertification, climate change \cite{Kougioumoutzis2020} and habitat loss \cite{ISPIKOUDIS1993259},
elevation gradient \cite{trigas2013, FAZAN2017}, grasing impacts on ecosystems \cite{JouffroyBapicot2016} and evolutionary model \cite{POULAKAKIS2002}.
Yet the terrestrial microbial diversity has been unexplored.

Has been sampled from major worldwide projects of microbiome studies \cite{Vasar2022, Labouyrie2023, Bahram2018, Orgiazzi2018}. Soil fungi \cite{Mikryukov2023, Davison2021, Tedersoo2021} and other soil eukaryotes \cite{Aslani2022}

\cite{Vogiatzakis2008_crete}
\cite{Vogiatzakis2016}

\cite{Sfenthourakis2017}
\cite{Vogiatzakis2008_med}

Integration of prior knowledge and 
Genome Standards Consortium \cite{Field2011}, FAIR data \cite{Wilkinson2016}, unrestricted \cite{Amann2019},
rich metadata and platforms of hosting open data like European Soil Data Centre \cite{Panagos2022}

Interactomes as a driver of community composition. Microbial interactions challenges \cite{Faust2021} cohesion?

@@

the data for this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB21776
ENA API \cite{Yuan2023}

@@

Copernicus CORINE Land Cover \cite{CLC2023}
WorldClim 2.0 \cite{Fick2017}
World Database on Protected Areas (WDPA) \cite{Hanson2022}
Function annotation - FAPROTAX \cite{Louca2016}
Crete geology from IGME
Aridity
%NDVI
Taxonomy Silva 138 \cite{Quast2012}

\subsection{Tools}\label{tools}
dada2 for ASV inferrence \cite{Callahan2016}
PEMA for OTU inferrence \cite{Zafeiropoulos2020}
ape R packege \cite{Paradis2004}
SRS R package \cite{Beule2020}
UMAP ordination \cite{mcinnes2018umap-software}
Numerical ecology analyses with vegan R package \cite{vegan}

Python 3.11.4
Julia Version 1.9.3 \cite{Julia-2017}
R version 4.3.2 \cite{rcoreteam}
Network inferrence using FlashWeave 0.19.2 \cite{Tackmann2019}
Network analysis in igraph R package \cite{Csardi2006}
Differential abundance analysis with ANCOM-BC R package \cite{Lin2020,Lin2023}
Spatial analysis with the sf and stars R packages \cite{Pebesma2023}
U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}
ggplot2 \cite{ggplot22016}
pheatmap \cite{Kolde2019}

Computations were performed on HPC infrastrure of HCMR \cite{Zafeiropoulos2021}.

@@

Networks specialists generalists \cite{Barberan2012}

Elevation gradients of biodiversity are known since Humboldt's work \cite{Rahbek2019} 
yet these patterns remain elusive regarding the soil microbiome \cite{Looby2020, Siles2023}.
Mostly because it's difficult to isolate other cofounding effects \cite{Nottingham2018}.

@@

Enumeratic all biodivarsity \cite{Anthony2023} and decipher the biogeochemical 
processes and interactions of fauna, flora and microbes \cite{Fry2019, Crowther2019,GRANDY201640,Delgado-Baquerizo2020}

Global hotspots \cite{Guerra2022}
and soil conservation instead of specific species \cite{Guerra2021}
and policy \cite{KONINGER2022} across countries \cite{Putten2023}

@@

Mountain peaks soil microbes \cite{Adamczyk2019}
Even though amplicon studies in soil should be interpeted with caution \cite{alteio2021} they 
can act as early warning signals towards public health conserns \cite{Banerjee2023}.
Some Conclusions here.
